A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-08-12 doc/large_strain.tex
Use updated images and fix minor typos

Note that the SIMD vectorization
in \texttt{deal.II} \cite{dealII90}
is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements{\color{red}, as opposed to the operations within an element}\footnote{%
In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
This motivates the caching strategy employed in Algorithm 3, as the pre-caching of the chosen data ensures a context in which SIMD parallelization of subsequent operations is guaranteed.
Algorithms 1 and 2 sacrifice some of this generality for decreased memory requirements.
}.
For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{%
For the heterogeneous materials considered here,
the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
}.
For example, the deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second-order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
In particular, \texttt{deal.II} provides the generic templated class \textit{VectorizedArray} that defines a unified interface to a vectorized data type
and overloads basic arithmetic operations based on compiler intrinsics.
The core of the sum factorization algorithms is provided by the
\textit{FEEvaluation} class, that supports all the necessary contractions such as $\overline{\grad} \gz N_i : (\boldsymbol{\mathcal{C}}\gz g_s  + \boldsymbol{\mathcal{G}} \gz g)$ to implement the three algorithms proposed above.
Thanks to the object-oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the locally owned elements.
For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well as SIMD vectorization) and performance tests, see \cite{kronbichler12}.

@@

The second implementation utilizes a matrix-free approach in conjunction with a geometric-multigrid (GMG) preconditioner.
GMG preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes.
For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
%
%
In this work, we adopt the matrix-free level operators within the GMG.
In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial. One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}.
Such operators are, however, computationally very demanding.
In this work we avoid these complications by assuming
that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material.
In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
We define the level operators as follows: Recall that the fine scale tangent operator \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$. We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
Essentially, level operators correspond to the linearization around the smoothed representation of displacement field.
Note that this differs from the classical approach where the (tangent) operator $\gz A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $\gz A^{l}$ on level $l$ via $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, where $\gz I^l_{l+1}$ and $\gz I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{CSL_Munich_throughput2d.eps}
      \caption{throughput matrix-vector product (2D)}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{CSL_Munich_timing2d.eps}
      \caption{time matrix-vector product (2D)}
      \label{fig:benchmark_miehe_CSL_vmult2}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{CSL_Munich_memory2d.eps}
      \caption{memory access (2D)}
      \label{fig:benchmark_miehe_CSL_memory2}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_throughput3d.eps}
    \caption{throughput matrix-vector product (3D)}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_timing3d.eps}
    \caption{time matrix-vector product (3D)}
    \label{fig:benchmark_miehe_CSL_vmult3}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_memory3d.eps}
    \caption{memory access (3D)}
    \label{fig:benchmark_miehe_CSL_memory3}
  \end{subfigure}
  \caption{Cascade Lake, one socket with 20 cores, matrix-vector multiplication.}%
  \label{fig:benchmark_miehe_CSL}
\end{figure}

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{IWR_newest_patched_throughput2d.eps}
      \caption{throughput matrix-vector product (2D)}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{IWR_newest_patched_timing2d.eps}
      \caption{time matrix-vector product (2D)}
      \label{fig:benchmark_miehe_IWR_vmult2}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{IWR_newest_patched_memory2d.eps}
      \caption{memory access (2D)}
      \label{fig:benchmark_miehe_IWR_memory2}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_throughput3d.eps}
    \caption{throughput matrix-vector product (3D)}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_timing3d.eps}
    \caption{time matrix-vector product (3D)}
    \label{fig:benchmark_miehe_IWR_vmult3}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_memory3d.eps}
    \caption{memory access (3D)}
    \label{fig:benchmark_miehe_IWR_memory3}
  \end{subfigure}
  \caption{IWR cluster, 20 cores, matrix-vector multiplication}%
  \label{fig:benchmark_miehe_IWR}
\end{figure}

@@

{\color{red}
This formulation can also be straight forwardly adopted within the matrix-free framework of \texttt{deal.II} library.
To that end a variation of the Algorithm \ref{alg:mf_tensor4} can be used where the gradients are evaluated with respect to the undeformed configuration and only a fourth-order referential tangent tensor ${\rm D}_{\gz F}{\gz P}$ is cached per quadrature point.
}

>>>>

Note that the SIMD vectorization
in \texttt{deal.II} \cite{dealII90}
is applied at the finite element level, i.e. the matrix-free operator is applied simultaneously on several elements (called ``blocks'').
The main reason is that, apart from the point-wise computation of the stresses and elastic tangents, the operations are typically the same on all elements{\color{red}, as opposed to the operations within an element}\footnote{%
In general, not all quadrature points in each element group are endowed with identical, non-dissipative constitutive laws that follow the same code path during evaluation or the kinetic quantities and tangents.
This motivates the caching strategy employed in Algorithm 3, as the pre-caching of the chosen data ensures a context in which SIMD parallelization of subsequent operations is guaranteed.
Algorithms 1 and 2 sacrifice some of this generality for decreased memory requirements.
}.
For a given quadrature point, we simultaneously perform tensor evaluations and contractions on all elements\footnote{%
For the heterogeneous materials considered here,
the elements are first grouped according to the material type (matrix or inclusion) and then the SIMD vectorization is applied for each group of elements.
This functionality is provided by the \texttt{deal.II} library in the \textit{MatrixFree} class.
}.
For example, the deformation gradient $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ (second-order tensor) is evaluated at the specific quadrature point of all elements within the ``block''.
This is achieved using templated number types within the \textit{Tensor} class of the \texttt{deal.II} library.
In particular, \texttt{deal.II} provides the generic templated class \textit{VectorizedArray} that defines a unified interface to a vectorized data type
and overloads basic arithmetic operations based on compiler intrinsics.
The core of the sum factorization algorithms is provided by the
\textit{FEEvaluation} class, that supports all the necessary contractions such as $\overline{\grad} \, \gz N_i : (\boldsymbol{\mathcal{C}}\gz g_s  + \boldsymbol{\mathcal{G}} \gz g)$ to implement the three algorithms proposed above.
Thanks to the object-oriented design of the \texttt{deal.II} library, the core of those matrix-free algorithms can be implemented in only a handful of lines.
Distributed memory MPI parallelization of the matrix-free operators is done using the standard domain decomposition technique, where each MPI process is responsible for applying the operator only on the locally owned elements.
For more information about the implementation details and data structures for sum factorization approaches in \texttt{deal.II} (including hanging constraints, storage of the inverse Jacobian of the transformation from unit to real cell, MPI and distributed memory parallelization as well as SIMD vectorization) and performance tests, see \cite{kronbichler12}.

@@

The second implementation utilizes a matrix-free approach in conjunction with a geometric-multigrid (GMG) preconditioner.
GMG preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes.
For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
%
%
In this work, we adopt the matrix-free level operators within the GMG.
In general, given a heterogeneous non-linear material, the definition of transfer operators\footnote{Finite dimensional linear operators from a vector space associated with the fine-scale mesh to the coarse-scale mesh, and in the opposite direction.} (for restriction and prolongation) and level operators is not trivial. One approach is to start from an arbitrary heterogeneous material at the fine scale and employ homogenization theory \cite{Suquet1987, Hill1972,Hashin1983,Castaneda1997} to design suitable transfer operators \cite{Miehe2007}.
Such operators are, however, computationally very demanding.
In this work we avoid these complications by assuming
that the mesh at the coarsest level can accurately describe the heterogeneous finite-strain elastic material.
In this case, the restriction and prolongation operations for patches of elements are always performed for the same material and thus admits usage of standard geometric transfer operators.
We define the level operators as follows: Recall that the fine scale tangent operator \eqref{eq:algebraic_tangent} is obtained by linearization around the current displacement field $\gz u^h$. We then restrict this displacement field to all multigrid levels $\{l\}$ and evaluate tangent operators using the matrix-free algorithms from Section \ref{sec:mf}.
Essentially, level operators correspond to the linearization around the smoothed representation of the displacement field.
Note that this differs from the classical approach where the (tangent) operator $\gz A^{l+1}$ on level $l+1$ is directly related to the (tangent) operator $\gz A^{l}$ on level $l$ via $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, where $\gz I^l_{l+1}$ and $\gz I^{l+1}_l$ are global prolongation (coarse-to-fine) and restriction (fine-to-coarse) operators, respectively.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{CSL_Munich_throughput2d-nolables.eps}
      \caption{throughput matrix-vector product (2D)}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{CSL_Munich_timing2d.eps}
      \caption{time matrix-vector product (2D)}
      \label{fig:benchmark_miehe_CSL_vmult2}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{CSL_Munich_memory2d-nolables.eps}
      \caption{memory access (2D)}
      \label{fig:benchmark_miehe_CSL_memory2}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_throughput3d-nolables.eps}
    \caption{throughput matrix-vector product (3D)}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_timing3d.eps}
    \caption{time matrix-vector product (3D)}
    \label{fig:benchmark_miehe_CSL_vmult3}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_memory3d-nolables.eps}
    \caption{memory access (3D)}
    \label{fig:benchmark_miehe_CSL_memory3}
  \end{subfigure}
  \caption{Cascade Lake, one socket with 20 cores, matrix-vector multiplication.}%
  \label{fig:benchmark_miehe_CSL}
\end{figure}

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{IWR_newest_patched_throughput2d-nolables.eps}
      \caption{throughput matrix-vector product (2D)}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{IWR_newest_patched_timing2d.eps}
      \caption{time matrix-vector product (2D)}
      \label{fig:benchmark_miehe_IWR_vmult2}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{IWR_newest_patched_memory2d-nolables.eps}
      \caption{memory access (2D)}
      \label{fig:benchmark_miehe_IWR_memory2}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_throughput3d-nolables.eps}
    \caption{throughput matrix-vector product (3D)}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_timing3d.eps}
    \caption{time matrix-vector product (3D)}
    \label{fig:benchmark_miehe_IWR_vmult3}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_memory3d-nolables.eps}
    \caption{memory access (3D)}
    \label{fig:benchmark_miehe_IWR_memory3}
  \end{subfigure}
  \caption{IWR cluster, 20 cores, matrix-vector multiplication}%
  \label{fig:benchmark_miehe_IWR}
\end{figure}

@@

{\color{red}
This formulation can also be straight forwardly adopted within the matrix-free framework of the \texttt{deal.II} library.
To that end a variation of the Algorithm \ref{alg:mf_tensor4} can be used where the gradients are evaluated with respect to the undeformed configuration and only a fourth-order referential tangent tensor ${\rm D}_{\gz F}{\gz P}$ is cached per quadrature point.
}
